---
title: "Analyses"
author: "Anders Skyrud Danielsen"
output:
  html_document: default
---

```{r setup and variables, include=FALSE}

# Set up R environment and define global chunk options
knitr::opts_chunk$set(echo = F, warning = F, message = F) # Disables echoing, warnings and messages in knitted output

# Load necessary packages
pacman::p_load(
  tidyverse,  # For data manipulation and visualization
  markdown,   # For markdown formatting
  gtsummary,  # For descriptive statistics
  broom,      # For tidying model outputs
  ggplot2,    # For data visualization
  viridis,    # For color palettes
  lubridate,  # For date manipulation
  rio,        # For data import/export
  scales,     # For graphical scales
  lmtest,     # For linear regression testing
  dotwhisker, # For regression visualization
  sandwich,   # For robust covariance matrix estimation
  incidence2, # For incidence computation
  knitr,      # For dynamic report generation
  splines     # For spline functions
)

# Set the default ggplot2 theme
theme_set(theme_classic())

# Set the working directory
setwd("~")

# Import the data
study1_data <- import("study1_data.rds")

# Data manipulation: Create new variables 'startdato' and 'sluttdato' and recode 'fodeland_string'
study1_data <- study1_data %>%
  mutate(startdato = ymd(DIAGNOSEDATO) - years(1),
         sluttdato = ymd(DIAGNOSEDATO) + years(3),
         sluttdato = if_else(notifiable==1, Proevedato, sluttdato),
         startslutt = sluttdato-startdato) %>%
  mutate(fodeland_2 = case_when(
    fodeland_string == "Norway" ~ "Norway",
    TRUE ~ "Other (or missing)"
  ))

```

# The epidemiology of multidrug-resistant organisms in cancer patients in Norway, 2008-2018 

## Analyses

These are the analyses as described in the manuscript. Please see the manuscript for introduction, methods, results, and discussion.

### Table 1

*Table 1. Characteristics of patients diagnosed with cancer in Norway from 2008 to 2018 at the time of their first cancer diagnosis, stratified by whether they were reported with multidrug-resistant organisms (MDROs) from six months prior to or up to three years after cancer diagnosis. MDROs include methicillin-resistant Staphylococcus aureus (MRSA), vancomycin- and/or linezolid-resistant enterococci (V/LRE), and carbapenemase-producing Gram-negative bacteria (CP-GNB).*

```{r table 1}

# Adjust variable levels and recode for the study1_data dataframe
study1_data <- study1_data %>% # English value names and reordering of levels
  mutate(alder_kat = factor(alder_kat, levels = c("<40 years", "40-49 years", "50-59 years", "60-69 years", 
                                                  "70-79 years", ">79 years"))) %>% # Reorder age categories
  mutate(inf_diag = factor(inf_diag, levels = c("MRSA", "VRE(/LRE)", "CPO"))) %>% # Reorder infection diagnosis categories
  mutate(KJOENN = recode(KJOENN, "K" = "F")) %>% # Recode gender variable
  mutate(Storby = case_when(fylke_string == "Oslo" ~ "Oslo", 
                            TRUE ~ "Other residency")) %>% # Recode residency variable
  mutate(forverring_string = case_when(forverring == 1 ~ "New cancer, same site",
                                       forverring == 2 ~ "New cancer, new site",
                                       forverring == 3 ~ "New metastasis",
                                        TRUE ~ "No new cancer")) # Recode cancer progression variable

# Generate summary table based on different microbe types
table1_a <- study1_data %>%
  select(ALDER, KJOENN, Storby, fodeland_2, diagnose_3, SEER_STADIUM, forverring_string, surgery, radiation, inf_diag) %>%
  tbl_summary(
    by = inf_diag,
    percent = "row",
    digits = all_categorical() ~ c(0,2),
    statistic = all_categorical() ~ "{n}",
    label = list(
      ALDER ~ "Age (continuous)", 
      #alder_kat ~ "Age (categorical)", 
      KJOENN ~ "Sex", 
      Storby ~ "Oslo", 
      fodeland_2 ~ "Country of birth", 
      diagnose_3 ~ "Type of cancer", 
      SEER_STADIUM ~ "Cancer dissemination",
      forverring_string ~ "Cancer progression",
      surgery ~ "Surgery reported", 
      radiation ~ "Radiotherapy reported"
    )
  ) %>%
  modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Resistant notifiable microbe**") %>%
  modify_header(
    stat_1 ~ "MRSA, n = {n}",
    stat_2 ~ "VRE(/LRE), n = {n}",
    stat_3 ~ "CPO, n = {n}",
  )

# Generate summary table based on notifiable disease status
table1_b <- study1_data %>%
  select(ALDER, KJOENN, Storby, fodeland_2, diagnose_3, SEER_STADIUM, forverring_string, surgery, radiation, notifiable) %>%
  tbl_summary(
    by = notifiable,
    percent = "row",
    digits = all_categorical() ~ c(0,2), 
    label = list(
      ALDER ~ "Age (continuous)", 
      #alder_kat ~ "Age (categorical)", 
      KJOENN ~ "Sex", 
      Storby ~ "Oslo", 
      fodeland_2 ~ "Country of birth", 
      diagnose_3 ~ "Type of cancer", 
      SEER_STADIUM ~ "Cancer dissemination",
      forverring_string ~ "Cancer progression",
      surgery ~ "Surgery reported",  
      radiation ~ "Radiotherapy reported"
    )
  ) %>%
  add_overall() %>%
  modify_header(
    stat_1 ~ "No notifiable microbe, n = {n}",
    stat_2 ~ "Notifiable microbe, n = {n}"
  )

# Merge two tables
table_1 <- tbl_merge(tbls = list(table1_b, table1_a), tab_spanner = F)

# Print the merged table
table_1

```

### Table 2

*Table 2. Characteristics of multidrug-resistant organisms (MDROs) in cancer patients six months prior to or up to three years after their first cancer diagnosis in Norway from 2008 to 2018, stratified by bacteria. MDROs include methicillin-resistant Staphylococcus aureus (MRSA), vancomycin- and/or linezolid-resistant enterococci (V/LRE), and carbapenemase-producing Gram-negative bacilli (CP-GNB).*


```{r table 2} 

# Update the study1_data dataframe with new variables
study1_data <- study1_data %>%
  mutate(inf_status_string = case_when(
    inf_status == 1 ~ "Infection", # Recode infection status as "Infection" when inf_status is 1
    inf_status == 0 ~ "Carriage", # Recode infection status as "Carriage" when inf_status is 0
    inf_status == 3 | (is.na(inf_status) & !is.na(notifiable)) ~ "Unknown" # Recode infection status as "Unknown" when inf_status is 3 or when inf_status is NA but notifiable is not NA
  )) %>%
  mutate(SmittetIUtlandet = case_when(
    SmittetIUtlandet == "Norge" ~ "Norway", # Recode SmittetIUtlandet as "Norway" when its value is "Norge"
    SmittetIUtlandet == "Utlandet" ~ "Abroad", # Recode SmittetIUtlandet as "Abroad" when its value is "Utlandet"
    SmittetIUtlandet == "Ukjent" | (is.na(Infeksjonsstatus) & !is.na(notifiable)) ~ "Unknown" # Recode SmittetIUtlandet as "Unknown" when its value is "Ukjent" or when Infeksjonsstatus is NA but notifiable is not NA
  ))

# Create a summary table using infection status, place of transmission, and infection diagnosis
table_2 <- study1_data %>% 
  select(inf_status_string, SmittetIUtlandet, inf_diag) %>%
  tbl_summary(
    by = inf_diag, # Group by infection diagnosis
    label = list(
      inf_status_string ~ "Infection status", # Set label for inf_status_string
      SmittetIUtlandet ~ "Place of transmission" # Set label for SmittetIUtlandet
    )
  ) %>% 
  add_overall() # Add overall statistics to the table

# Print table_2
table_2

```

### Figure 2

*Figure 2. Incidence rates of infection, colonisation and total of multidrug-resistant organisms (MDROs) by person-years, found in cancer patients six months prior to or up to three years after their first cancer diagnosis in Norway from 2008 to 2018. Dashed lines represent observation time outside of the inclusion period, with diminishing person-years. MDROs include methicillin-resistant Staphylococcus aureus (MRSA), vancomycin- and/or linezolid-resistant enterococci (VRE/LRE), and carbapenemase-producing Gram-negative bacilli (CP-GNB).*

```{r figure 2, fig.height=12} 

# Calculate the incidence of different infection statuses and diagnoses per year
incidence <- incidence(study1_data, Proevedato, interval = "years", groups = c(inf_diag, inf_status_2)) 
incidence$date_index <- as.double(incidence$date_index)

# Add rows for missing years and infection status/diagnosis combinations
incidence <- incidence %>%
  add_row(date_index = as.double(2007),
          inf_diag = "MRSA",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2007),
          inf_diag = "MRSA",
          inf_status_2 = 1,
          count = 0) %>%
  add_row(date_index = as.double(2007),
          inf_diag = "VRE(/LRE)",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2007),
          inf_diag = "VRE(/LRE)",
          inf_status_2 = 1,
          count = 0) %>%
  add_row(date_index = as.double(2008),
          inf_diag = "VRE(/LRE)",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2008),
          inf_diag = "VRE(/LRE)",
          inf_status_2 = 1,
          count = 0) %>%
  add_row(date_index = as.double(2009),
          inf_diag = "VRE(/LRE)",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2012),
          inf_diag = "CPO",
          inf_status_2 = 1,
          count = 0) %>%
  add_row(date_index = as.double(2013),
          inf_diag = "CPO",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2015),
          inf_diag = "CPO",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2016),
          inf_diag = "CPO",
          inf_status_2 = 1,
          count = 0) %>%
  add_row(date_index = as.double(2021),
          inf_diag = "MRSA",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2021),
          inf_diag = "VRE(/LRE)",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2021),
          inf_diag = "VRE(/LRE)",
          inf_status_2 = 1,
          count = 0) %>%
  add_row(date_index = as.double(2021),
          inf_diag = "CPO",
          inf_status_2 = 0,
          count = 0) %>%
  add_row(date_index = as.double(2021),
          inf_diag = "CPO",
          inf_status_2 = 1,
          count = 0) %>%
  group_by(inf_diag, date_index) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                      inf_status_2=2))

# Create a data frame with the start and end dates of each year
df_years <- tibble(
  yr=2007:2021
) %>% 
  mutate(yr_start = paste0(yr,'-01-01'), yr_end = paste0(yr, '-12-31')) %>% 
  type_convert()

# Create a data frame with the number of days each patient contributed to each year
df_out <- study1_data %>% 
  crossing(df_years) %>% 
  mutate(days_in_yr = pmin(sluttdato, yr_end) - pmax(startdato, yr_start)) %>% 
  filter(days_in_yr>0)

# Sum the person-days for each year
person_days <- df_out %>% 
  group_by(yr) %>% 
  summarise(person_days=sum(days_in_yr))

# Merge the person_days data frame with the incidence data frame and calculate the incidence rate
incidence <- incidence %>%
  left_join(person_days, by = c("date_index" = "yr")) %>%
  mutate(personyears = person_days/365.25) %>%
  mutate(incidence_rate = (count/as.double(personyears))*100000) 

# Fit Poisson regression models to the data
Poisson.MRSArate <- glm(count ~ date_index + offset(log(as.numeric(personyears))), family = "poisson", data = subset(incidence, inf_diag == "MRSA" & inf_status_2 == 2))
Poisson.VRErate <- glm(count ~ date_index + offset(log(as.numeric(personyears))), family = "poisson", data = subset(incidence, inf_diag == "VRE(/LRE)" & inf_status_2 == 2))
Poisson.CPOrate <- glm(count ~ date_index + offset(log(as.numeric(personyears))), family = "poisson", data = subset(incidence, inf_diag == "CPO" & inf_status_2 == 2))

# Print summaries of the models
print(summary(Poisson.MRSArate))
print(summary(Poisson.VRErate))
print(summary(Poisson.CPOrate))

# Create line plots of the incidence rates over time
ggplot_MRSA <- ggplot(data = subset(incidence, inf_diag == "MRSA"), aes(y=incidence_rate, x=date_index, group = inf_status_2)) +
  geom_line(data=subset(incidence, inf_diag == "MRSA" & date_index <= 2008), linetype = "dashed", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_line(data=subset(incidence, inf_diag == "MRSA" & date_index >= 2018), linetype = "dashed", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_line(data=subset(incidence, inf_diag == "MRSA" & (date_index >= 2008 & date_index <= 2019)), linetype = "solid", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_point(aes(colour = factor(inf_status_2), alpha = as.numeric(personyears)), size = 2, shape = 19) +
  ylab("Incidence rate per 100,000 person-years") +
  xlab("Year of follow-up time") +
  ggtitle("MRSA") +
  scale_x_continuous(n.breaks = 14) +
  theme_classic(base_size = 12) + 
  scale_colour_manual(name = "Infection status", values = c("#978bc4", "#523f9e", "#31255e"), labels = c("Colonisation or unknown", "Infection", "Total")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(alpha = F)

ggplot_VRE <- ggplot(data = subset(incidence, inf_diag == "VRE(/LRE)"), aes(y=incidence_rate, x=date_index, group = inf_status_2)) +
  geom_line(data=subset(incidence, inf_diag == "VRE(/LRE)" & date_index <= 2008), linetype = "dashed", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_line(data=subset(incidence, inf_diag == "VRE(/LRE)" & date_index >= 2018), linetype = "dashed", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_line(data=subset(incidence, inf_diag == "VRE(/LRE)" & (date_index >= 2008 & date_index <= 2019)), linetype = "solid", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_point(aes(colour = factor(inf_status_2), alpha = as.numeric(personyears)), size = 2, shape = 19) +
  ylab("Incidence rate per 100,000 person-years") +
  xlab("Year of follow-up time") +
  ggtitle("VRE(/LRE)") +
  scale_x_continuous(n.breaks = 14) +
  theme_classic(base_size = 12) + 
  scale_colour_manual(name = "Infection status", values = c("#978bc4", "#523f9e", "#31255e"), labels = c("Colonisation or unknown", "Infection", "Total")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(alpha = F)

ggplot_CPO <- ggplot(data = subset(incidence, inf_diag == "CPO"), aes(y=incidence_rate, x=date_index, group = inf_status_2)) +
  geom_line(data=subset(incidence, inf_diag == "CPO" & (date_index < 2008 | date_index > 2018)), linetype = "dashed", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_line(data=subset(incidence, inf_diag == "CPO" & (date_index >= 2008 & date_index <= 2019)), linetype = "solid", aes(colour = factor(inf_status_2)), size=1.2) +
  geom_point(aes(colour = factor(inf_status_2), alpha = as.numeric(personyears)), size = 2, shape = 19) +
  # geom_vline(xintercept = 2012, linetype = "dashed", size = 1.5, alpha = 0.6) +
  ylab("Incidence rate per 100,000 person-years") +
  xlab("Year of follow-up time") +
  ggtitle("CP-GNB") +
  scale_x_continuous(n.breaks = 14) +
  # annotate(x=2013, y=9.5, label = "CPO \n became \n notifiable", geom = "label") +
  theme_classic(base_size = 12) + 
  scale_colour_manual(name = "Infection status", values = c("#978bc4", "#523f9e", "#31255e"), labels = c("Colonisation or unknown", "Infection", "Total")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(alpha = F)

# Arrange the plots into a single figure
figur2 <- ggarrange(ggplot_MRSA, ggplot_VRE, ggplot_CPO, ncol = 1, common.legend = T, heights = 2) 

# Print the figure
figur2

# Create a table of person-years for each year
incidence_table <- incidence[,c("date_index", "personyears")]
incidence_table <- incidence_table %>%
  mutate(personyears = as.numeric(round(personyears, 0))) %>%
  distinct(date_index, personyears) %>%
  rename(Personyears = personyears)

```

### Table S2

*Table S2. Person-years included in the incidence rate calculation in figure 2.*

```{r table S2}

kable(incidence_table, col.names = c("Year", "Person-years"))

```

### Figure 3

*Figure 3. Density estimate for the finding of multidrug-resistant organisms (MDROs) by time since cancer diagnosis six months prior to or up to three years after the first cancer diagnosis of persons in Norway, from 2008 to 2018. MDROs include methicillin-resistant Staphylococcus aureus (MRSA), vancomycin- and/or linezolid-resistant enterococci (V/LRE), carbapenemase-producing Gram-negative bacilli (CP-GNB), and third-generation cephalosporin-resistant Enterobacterales (3GCR-E).*

```{r figur 3}
# Calculate time until infection based on either notifiable status or TGCRE status
study1_data_tid <- study1_data %>% 
  mutate(tid_til_smitte = case_when( 
    notifiable == 1 ~ ymd(Proevedato)-ymd(startdato),
    TGCRE == 1 ~ ymd(SampleDate)-ymd(startdato))) %>% 
  mutate(tid_til_smitte = as.numeric(tid_til_smitte)) %>%
  mutate(inf_diag = case_when(
    TGCRE == 1 ~ "3GCR-E",
    inf_diag == "MRSA" ~ "MRSA",
    inf_diag == "VRE(/LRE)" ~ "V/LRE",
    inf_diag == "CPO" ~ "CP-GNB"
  )) %>% 
  filter(tid_til_smitte<1500) %>%
  mutate(tid_til_smitte = tid_til_smitte-365)

# Create histogram of time until infection
tid_til_smitte <- ggplot(data = study1_data_tid) +
  geom_histogram(mapping = aes(x = tid_til_smitte, fill = inf_diag), binwidth = 122) +
  scale_fill_viridis_d(option = "D", end = 0.9, alpha = 0.8) +
  guides(fill = guide_legend("Resistant bacteria")) +
  scale_x_continuous(n.breaks = 10) +
  ylab("Counts") +
  xlab("Days since diagnosis") 

# Create density plot of time until infection
tid_til_smitte <- ggplot(data = study1_data_tid) +
  geom_density(mapping = aes(x = tid_til_smitte, colour = inf_diag), size = 1.5) +
  ylab("Density estimate") +
  xlab("Days since diagnosis") +
  geom_vline(xintercept = 0, linetype = "dashed", size = 1.5, alpha = 0.6) +
  annotate(x=0, y=0.0035, label = "Cancer diagnosis", geom = "label") +
  theme_classic(base_size = 12) + 
  guides(colour = guide_legend("Resistant bacteria")) +
  scale_x_continuous(n.breaks = 12) +
  scale_color_viridis_d(option = "D", direction = -1, end = 0.9, alpha = 0.8)

# Return the density plot
tid_til_smitte

```

### Figure 4

*Figure 4. Age- and sex-standardised incidence rate ratios (IRR) of infection/colonisation with either methicillin-resistant Staphylococcus aureus (MRSA) or vancomycin- and/or linezolid-resistant enterococci (V/LRE) in cancer patients six months prior to or up to three years after their first cancer diagnosis in Norway, from 2008 to 2018, by different cancer types.*

```{r figur 4}

# Regression

# Recode the data and calculate the diagnosis year
study1_data <- study1_data %>%
  mutate(MRSA = ifelse(inf_diag == "MRSA", 1, NA)) %>% 
  mutate(VRE = ifelse(inf_diag == "VRE(/LRE)", 1, NA)) %>% 
  replace_na(list(MRSA = 0, VRE = 0)) %>% 
  mutate(diagnose_3 = fct_relevel(diagnose_3, "Gastrointestinal")) %>% 
  mutate(diagnosisyear = year(DIAGNOSEDATO))

# Fit Poisson model for MRSA
Poisson.MRSA <- glm(MRSA ~ ALDER + KJOENN + diagnose_3 + offset(log(as.numeric(startslutt))), family = "poisson", data = study1_data)

# Use robust standard errors
Poisson.robust.MRSA <- coeftest(Poisson.MRSA,
                           vcov = sandwich)

# Prepare coefficients for presentation
Poisson.robust.MRSA <- tidy(Poisson.robust.MRSA, conf.int = TRUE, exponentiate = TRUE)

# Further processing of coefficients for MRSA
Poisson.robust.MRSA <- Poisson.robust.MRSA %>% 
  mutate(estimate = exp(estimate),
         conf.low = exp(conf.low),
         conf.high = exp(conf.high)) %>% 
  filter(!str_detect(term, "ALDER") & term!= "KJOENNM" & term!= "(Intercept)") %>% 
  mutate(term = str_sub(term, 11)) %>% 
  arrange(desc(estimate)) %>% 
  add_row(term = "Gastrointestinal", estimate = 1, .before = 1)

# Fit Poisson model for VRE
Poisson.VRE <- glm(VRE ~ ALDER + KJOENN + diagnose_3 + offset(log(as.numeric(startslutt))), family = "poisson", data = study1_data)

# Here is the code where I decided not to use splines
# I checked the BIC and performed a likelihood ratio test between the models
#Poisson.VREspline <- glm(VRE ~ ns(ALDER, 2) + KJOENN + diagnose_3 + ns(diagnosisyear, 5) + offset(log(as.numeric(startslutt))), family = "poisson", data = study1_data)
#Poisson.VRElinear <- glm(VRE ~ ALDER + KJOENN + diagnose_3 + offset(log(as.numeric(startslutt))), family = "poisson", data = study1_data)

#BIC(Poisson.VREspline)
#BIC(Poisson.VRElinear)

#lrtest(Poisson.VREspline, Poisson.VRElinear)

# This is slightly worse, as is restricted cubic splines
#ald_quants <- quantile(study1_data$ALDER, probs = c(0.33, 0.67))
#Poisson.VRE_quants <- glm(VRE ~ ns(ALDER, knots = ald_quants) + KJOENN + diagnose_3 + offset(log(as.numeric(startslutt))), family = "poisson", data = study1_data)
#AIC(Poisson.VRE_quants) 

# plot to check predicted vs observed for the spline
# Add the observed values of VRE and ALDER to the dataframe
#study1_data$Observed_VRE <- study1_data$VRE
#study1_data$ALDER <- study1_data$ALDER

#newdata <- tibble(KJOENN = "K", diagnose_3 = "Gastrointestinal", ALDER = seq(18,85,1), startslutt = 1461)

# Predict the values using the Poisson regression model
#newdata$Predicted_VRE <- predict(Poisson.VRE, newdata = newdata, type = "response")

# Calculate the mean predicted and observed VRE values for each unique ALDER value
#mean_values <- study1_data %>%
#  group_by(ALDER) %>%
#  summarise(Mean_Predicted_VRE = mean(Predicted_VRE),
#           Mean_Observed_VRE = mean(Observed_VRE))

# Create a scatterplot of mean observed vs mean predicted values
#ggplot(newdata, aes(x = ALDER)) +
#  geom_line(aes(y = Predicted_VRE), color = "red", size = 1) +
#  labs(x = "ALDER", y = "VRE") +
#  ggtitle("Mean Observed vs Mean Predicted Values of VRE") +
#  theme_minimal()

# Robust standard errors for VRE
Poisson.robust.VRE <- coeftest(Poisson.VRE,
                           vcov = sandwich)

# Prepare coefficients for presentation
Poisson.robust.VRE <- tidy(Poisson.robust.VRE, conf.int = TRUE, exponentiate = TRUE)

# Further processing of coefficients for VRE
Poisson.robust.VRE <- Poisson.robust.VRE %>% 
  mutate(estimate = exp(estimate),
         conf.low = exp(conf.low),
         conf.high = exp(conf.high)) %>% 
  filter(!str_detect(term, "ALDER") & term!= "KJOENNM" & term!= "(Intercept)") %>% 
  mutate(term = str_sub(term, 11)) %>% 
  arrange(desc(estimate)) %>% 
  add_row(term = "Gastrointestinal", estimate = 1, .before = 1)

# Prepare theme for plots
theme_set(theme_bw())

# Create the plot
plot.new()
par(c(2,1))
figure4_a <- dwplot(Poisson.robust.MRSA, dot_args = list(size = 3), whisker_args = list(size = 1)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 20)) +
  geom_vline(xintercept = 1) +
  xlab("Incidence rate ratio") +
  ylab("Cancer types") +
  labs(title = "MRSA") +
  theme_bw(base_size = 12) +
  scale_color_viridis_d(begin = 0.3)

# Create the second plot
figure4_b <- dwplot(Poisson.robust.VRE, dot_args = list(size = 3), whisker_args = list(size = 1)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 20)) +
  geom_vline(xintercept = 1) +
  xlab("Incidence rate ratio") +
  ylab("Cancer types") +
  labs(title = "V/LRE") +
  theme_bw(base_size = 12) +
  scale_color_viridis_d(begin = 0.3)

# Arrange and display the plots
figur4 <- ggarrange(figure4_a, figure4_b, ncol = 1, heights = 2, legend = "none")
figur4

# Fit Poisson model for healthcare workers
Poisson.healthcare <- glm(notifiable ~ ALDER + KJOENN + helse + offset(log(as.numeric(startslutt))), family = "poisson", data = study1_data)

# Robust standard errors for healthcare model
Poisson.robust.healthcare <- coeftest(Poisson.healthcare,
                           vcov = sandwich)

# Prepare coefficients for presentation
Poisson.robust.healthcare <- tidy(Poisson.robust.healthcare, conf.int = TRUE, exponentiate = TRUE)

```

### Table 3

*Table 3. Characteristics of patients diagnosed with or treated for cancer at Oslo University Hospital from 2008 to 2018, in whom Enterobacterales were found in blood or urine from six months prior to or up to three years after first cancer diagnosis. The table is stratified by whether they were found to have a third-generation cephalosporin-resistant Enterobacterales (3GCR-E) or third-generation cephalosporin-susceptible Enterobacterales (3GCS-E).*


```{r table 3}

# Recoding species names
study1_data <- study1_data %>% 
  mutate(Species = case_when(
    Species == "Andre Escherichia spp." ~ "Other findings",
    Species == "Andre funn" ~ "Other findings",
    TRUE ~ Species
  ))

# Creating summary table with select variables
table3 <- study1_data %>%
  select(ALDER, KJOENN, Storby, fodeland_2, Species, diagnose_3, SEER_STADIUM, forverring_string, surgery, radiation, TGCRE) %>%
  tbl_summary(
    by = TGCRE,
    percent = "row",
    digits = all_categorical() ~ c(0,2), 
    label = list(
      ALDER ~ "Age (continuous)", 
      #alder_kat ~ "Age (categorical)", 
      KJOENN ~ "Sex", 
      Storby ~ "Oslo", 
      fodeland_2 ~ "Country of birth", 
      Species ~ "Species",
      diagnose_3 ~ "Type of cancer", 
      SEER_STADIUM ~ "Cancer dissemination",
      forverring_string ~ "Cancer progression",
      surgery ~ "Surgery reported",
      radiation ~ "Radiotherapy reported"
    )
  ) %>%
  add_overall() %>%  # Add overall column
  modify_header(
    stat_1 ~ "3GCS-E, n = {n}",  # Modify header for first stats column
    stat_2 ~ "3GCR-E, n = {n}"  # Modify header for second stats column
  )

# Print the table
table3

``` 

### Figure 5

*Figure 5: The proportion of Enterobacterales resistant to third-generation cephalosporins found in the blood or urine of patients diagnosed with or treated for cancer at Oslo University Hospital by year the sample was taken, found from six months prior to or up to three years after the first cancer diagnosis from 2008 to 2018.*

```{r figure 5}
# Select only rows with non-NA TGCRE values and add a year column for the sample date
study1_data_figure5 <- study1_data %>% 
  mutate(TGCRE_select = ifelse(!is.na(TGCRE), 1, NA)) %>%
  filter(!is.na(TGCRE_select)) %>%
  arrange(desc(TGCRE)) %>%
  mutate(SampleYear = year(SampleDate))

# Calculate summary statistics and confidence intervals for each year
figure5_CI <- study1_data_figure5 %>%
  group_by(SampleYear) %>%
  summarise(n=n(), n_pos=sum(TGCRE), pct_pos = round(100*n_pos/n, 1)) %>% 
  rowwise() %>% 
  mutate(lci = (round(100*binom.test(n_pos, n)$conf.int[1],1))/100,
         uci = (round(100*binom.test(n_pos, n)$conf.int[2],1))/100
         ) 

# Merge the summary statistics back into the data
study1_data_figure5 <- study1_data_figure5 %>% 
  left_join(figure5_CI, by = "SampleYear")

# Create the plot
figure5 <- ggplot(study1_data_figure5, aes(SampleYear, y = TGCRE_select, fill = factor(TGCRE))) + 
  geom_col(position = "fill") + 
  geom_errorbar(aes(x=SampleYear, ymin=lci, ymax=uci, width=0.3)) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(name = "Enterobacterales", values = c("#978bc4", "#523f9e"), labels = c("Susceptible", "Resistant")) +
  coord_cartesian(ylim = c(0,0.25)) +
  ylab("Proportion of Enterobacterales resistant to \n third-generation cephalosporins") +
  xlab("Year of sample") +
  guides(colour = guide_legend("Enterobacterales")) +
  scale_x_continuous(breaks = c(2007,2009,2011,2013,2015,2017,2019,2021)) 

# Display the plot
figure5

# Calculate person-days for each year
df_out_TGCRE <- study1_data %>% 
  filter(!is.na(TGCRE)) %>% 
  crossing(df_years) %>% 
  mutate(days_in_yr = pmin(sluttdato, yr_end) - pmax(startdato, yr_start)) %>% 
  filter(days_in_yr>0)

person_days_TGCRE <- df_out_TGCRE %>% 
  group_by(yr) %>% 
  summarise(person_days=sum(days_in_yr))

# Merge person-days into the data
figure5_CI <- figure5_CI %>%
  left_join(person_days_TGCRE, by = c("SampleYear" = "yr")) %>%
  mutate(personyears = person_days/365.25) 

# Fit and print a Poisson regression model
Poisson.TGCRErate <- glm(n_pos ~ SampleYear + offset(log(as.numeric(personyears))), family = "poisson", data = figure5_CI)
print(summary(Poisson.TGCRErate))

```

### Export of figures

```{r eksport, eval=F}

# Exports

ggsave("figure2.png", plot = figur2, dpi = 320, width = 25, height = 50, units = "cm")
ggsave("figure3.png", plot = figur4, dpi = 320, width = 25, height = 30, units = "cm")
ggsave("figure4.png", plot = figure5, dpi = 320, width = 25, height = 20, units = "cm")
ggsave("figureS2.png", plot = tid_til_smitte, dpi = 320, width = 25, height = 25, units = "cm")


```

